
libraries
```{r}
library('prism')
library('ggplot2')
library('ggeffects')
library('dplyr')
library('raster')
library('lubridate')
library('scales')
library('mgcv')
library('gratia')
library('pROC')
library('tidycensus')
```

prism directory
```{r}
prism_set_dl_dir('/Users/osinachinwosu/Desktop/TMAX')
```

loading lead_data
```{r}
lead_data = read.csv('/Users/osinachinwosu/Desktop/JHU_Intern/leadExposure_final.csv')
```

```{r}
lead_copy = lead_data
unique_dates = unique(lead_copy$Date.Sampled) #12234 unique dates
lead_copy
```

```{r}
lead_copy$Date = lapply(lead_copy$Date.Sampled, as.Date)
lead_copy$Date = as.Date(unlist(lead_copy$Date))
```

creating table + merging 
```{r}
mx_df = data.frame(Date.Sampled = character(), cxy_lon = double(), cxy_lat = double(), tmax = double(), first_draw = double(), second_draw = double(), third_draw = double(), Address = character(), cxy_block_id = integer())

for (i in 1:nrow(lead_copy)) {
    date2 = lead_copy$Date[i]
    proper_date = format(date2, '%Y%m%d')
    longitude = lead_copy$cxy_lon[i]
    latitude = lead_copy$cxy_lat[i]
    draw_1 = lead_copy$X1st.Draw[i]
    draw_2 = lead_copy$X2.3.Minute[i]
    draw_3 = lead_copy$X5.Minute[i]
    address = lead_copy$Address[i]
    block = lead_copy$cxy_block_id[i]
    
    file_labels = prism_archive_ls()
    file_labels = file_labels[grep(proper_date, file_labels)] 
    if (length(file_labels) > 0) {
      prism_file = pd_to_file(file_labels)
      if (file.exists(prism_file)) {
        raster_file = raster(prism_file)
        temp_C = extract(raster_file, matrix(c(longitude, latitude), ncol = 2))
        temp = if (!is.na(temp_C)) {
          temp_C
              } else {
                NA
                }
mx_df = rbind(mx_df, data.frame(Date.Sampled = date2, cxy_lon = longitude, cxy_lat = latitude, tmax = temp, X1st.Draw = draw_1, X2.3.Minute = draw_2, X5.Minute = draw_3, Address = address, cxy_block_id = block))
        }
    }
}

```

```{r}
merged_mx = distinct(left_join(lead_copy, max_df, by = c("Address", "cxy_lon", "cxy_lat", "X1st.Draw", "X2.3.Minute", "X5.Minute", 'Date')))
lead_copy = distinct(lead_copy)
```

LOG LM'S + GRAPHS (Week 3 - 4)
```{r}
ggplot(merged_mx, aes(x = Date.Sampled, y = tmax)) +
  geom_line() +
  labs(title = "Temperature Over Time",
       x = "Date",
       y = "Average Temperature (°C)")
```

```{r}
# Outlier's Included
draw_temp_1 = ggplot(merged_mx, aes(x = tmax, y = X1st.Draw)) +
  geom_point() +
  labs(title = "Temperature vs 1st Draw Lead Concentration",
       x = "Average Temperature (°C)",
       y = "1st Draw Lead Concentration")

draw_temp_23 = ggplot(merged_mx, aes(x = tmax, y = X2.3.Minute)) +
  geom_point() +
  labs(title = "Temperature vs 2nd Draw Lead Concentration",
       x = "Average Temperature (°C)",
       y = "2nd Draw Lead Concentration After 2-3 Minutes")

draw_temp_5 = ggplot(merged_mx, aes(x = tmax, y = X5.Minute)) +
  geom_point() +
  labs(title = "Temperature vs 3rd Draw Lead Concentration",
       x = "Average Temperature (°C)",
       y = "3rd Draw Lead Concentration After 5 Minutes")


draw_temp_1
draw_temp_23
draw_temp_5
```

```{r}
#Outlier Removal Function
orm2 = function(df, column) {
  quartiles = quantile(df[[column]], probs = c(0.25,0.75), na.rm=TRUE)
  IQR = IQR(df[[column]])
  lower = quartiles[1] - 1.5*IQR
  upper = quartiles[2] + 1.5*IQR
  noout_df2 = subset(df, df[[column]] < upper)
  return(noout_df2)
}
```

```{r}
# Outlier's Excluded - non-log graphs
draw1_noouts = orm2(merged_mx, "X1st.Draw")
draw2_noouts = orm2(merged_mx, "X2.3.Minute")
draw3_noouts = orm2(merged_mx, "X5.Minute")

draw_temp_1_nl = ggplot(draw1_noouts, aes(x = tmax, y = X1st.Draw)) +
  geom_point() +
  labs(title = "Temperature vs 1st Draw Lead Concentration",
       x = "Average Temperature (°C)",
       y = "1st Draw Lead Concentration")

draw_temp_23_nl = ggplot(draw2_noouts, aes(x = tmax, y = X2.3.Minute)) +
  geom_point() +
  labs(title = "Temperature vs 2nd Draw Lead Concentration",
       x = "Average Temperature (°C)",
       y = "2nd Draw Lead Concentration After 2-3 Minutes")

draw_temp_5_nl = ggplot(draw3_noouts, aes(x = tmax, y = X5.Minute)) +
  geom_point() +
  labs(title = "Temperature vs 3rd Draw Lead Concentration",
       x = "Average Temperature (°C)",
       y = "3rd Draw Lead Concentration After 5 Minutes")


draw_temp_1_nl
draw_temp_23_nl
draw_temp_5_nl
```

lead vs temp graphs
```{r}
#Outlier's excluded - log graphs
g_draw1 =  ggplot(draw1_noouts, aes(x = log(tmax), y = X1st.Draw)) +
  geom_point(alpha = .5) +
  labs(title = "Temperature vs 1st Draw Lead Concentration (Removed Upper Quartile Outliers)",
       x = "Log Average Temperature (°C)",
       y = "1st Draw Lead Concentration")

g_draw23 = ggplot(draw2_noouts, aes(x = log(tmax), y = X2.3.Minute)) +
  geom_point(alpha = .5) +
  labs(title = "Temperature vs 2nd Draw Lead Concentration (Removed Upper Quartile Outliers)",
       x = "Log Average Temperature (°C)",
       y = "2nd Draw Lead Concentration After 2-3 Minutes")

g_draw5 = ggplot(draw3_noouts, aes(x = log(tmax), y = X5.Minute)) +
  geom_point(alpha = .5) +
  labs(title = "Temperature vs 3rd Draw Lead Concentration (Removed Upper Quartile Outliers)",
       x = "Log Average Temperature (°C)",
       y = "3rd Draw Lead Concentration After 5 Minutes")

g_draw1
g_draw23
g_draw5
```

Linear Models using temperature and communities areas on flush
```{r}
summary(lm(X1st.Draw ~ tmax + as.factor(CA), data = merged_mx))

summary(lm(X1st.Draw ~ tmax, data = merged_mx))
```

```{r}
summary(lm(X2.3.Minute ~ tmax + as.factor(CA), data = merged_mx))

summary(lm(X2.3.Minute ~ CA, data = merged_mx))
```

```{r}
summary(lm(X5.Minute ~ tmax + as.factor(CA), data = merged_mx))

summary(lm(X5.Minute ~ CA, data = merged_mx))
```

GAM Models and graphs using temperature and community area on flush (Week 4)

```{r}
mod_gam1 = gam(X1st.Draw ~ as.factor(CA) + s(tmax), data = merged_mx)
summary(mod_gam1)
gratia::draw(mod_gam1)
```

```{r}
mod_gam3 = gam(X2.3.Minute ~ as.factor(CA) + s(tmax), data = merged_mx)
summary(mod_gam3)
gratia::draw(mod_gam3)
```
```{r}
mod_gam2 = gam(X5.Minute ~ s(tmax) + as.factor(CA), data = merged_mx)
summary(mod_gam2)
gratia::draw(mod_gam2)
```

Log regression + look into subtopics (Week 5)
```{r}
#took the log of tmax after adding  a constant (min of temp + 1) to each row
merged_mx$log_temp = log(abs(min(merged_mx$tmax)) + merged_mx$tmax + 1)
```

```{r}
merged_mx = merged_mx %>% mutate(lead_presence = ifelse(X1st.Draw > 0, ifelse(X2.3.Minute > 0, ifelse(X5.Minute > 0, 1, 0), 0), 0))
merged_mx
```

regression
```{r}
log_model = glm(lead_presence ~ as.factor(CA) + log_temp, data = merged_mx, family = binomial)

summary(log_model)

non_significant = summary(log_model)$coefficients[summary(log_model)$coefficients[, "Pr(>|z|)"] > 0.05, ]
print(non_significant)
```


Week 5 + 6: Continuation of Neighborhood
```{r}
imputeDF = read.csv('/Users/osinachinwosu/Desktop/imputeDF.csv')
imputeDF
```

Week 7+8:

BenHuynh github
```{r}
riskDF = read.csv("/Users/osinachinwosu/Downloads/riskDF.csv")
imputeDF = read.csv("/Users/osinachinwosu/Desktop/imputeDF.csv")
imputeDF2 = imputeDF %>% select(blockPopulation,
                                             propWhiteBlockPop,
                                             propBlackBlockPop,
                                             propAianBlockPop,
                                             propAsianBlockPop,
                                             propHispanicBlockPop,
                                             Population,
                                             pctW,pctB,pctA,
                                             pctH,pctP,pctQ,
                                             whitePropBG,blackPropBG,
                                             aianPropBG,asianPropBG,
                                             hispanicPropBG,
                                             educationHSPropBG,educationGEDPropBG,
                                             educationAssociatesPropBG,
                                             educationBachelorsPropBG,
                                             educationSomeCollegeLess1PropBG,
                                             educationSomeCollegeMore1PropBG,
                                             educationMastersPropBG,
                                             educationPrfsnlSchoolPropBG,
                                             educationDoctoratePropBG,blockGroup,
                                 speakOtherLanguagesPropBG,pRenterOccupiedHousesBG, foreignBornPropBG,
                                             censusTract,blockNum,tested,overOne_2) %>%  distinct(blockNum,.keep_all=T)

#riskDF = riskDF %>% select(-overOne_2) %>% distinct()
riskDF$blockNum = as.character(riskDF$blockNum)
outcomeDF = imputeDF %>% filter(tested) %>%
  group_by(blockNum) %>% mutate(meanOutcome = mean(as.numeric(overOne_2)),
                                meanOutcome = ifelse(meanOutcome>=0.5,TRUE,FALSE)) %>% 
  ungroup() %>% distinct(blockNum,.keep_all=T)
imputeDF2$blockNum = as.character(imputeDF2$blockNum)
rarDF = imputeDF2 %>% left_join(riskDF,by=c("blockNum"="blockNum"))
outcomeTestsDF = imputeDF %>% filter(tested)
```

Outcome variable - meanOutcome

Used blockPopulation as control variable
```{r}
model_sol = glm(meanOutcome ~ speakOtherLanguagesPropBG + blockPopulation, data = outcomeDF, family = binomial) %>%summary()
model_roh = glm(meanOutcome ~ pRenterOccupiedHousesBG + blockPopulation, data = outcomeDF, family = binomial) %>%summary()
model_fb = glm(meanOutcome ~ foreignBornPropBG + blockPopulation, data = outcomeDF, family = binomial) %>%summary()

model_sol
model_roh
model_fb
```

Outcome Variable - Tested
```{r}
tstsol = glm(tested ~ speakOtherLanguagesPropBG + blockPopulation, data = rarDF, family = binomial) %>%summary()
tstroh = glm(tested ~ pRenterOccupiedHousesBG + blockPopulation, data = rarDF, family = binomial) %>%summary()
tstfb = glm(tested ~ foreignBornPropBG + blockPopulation, data = rarDF, family = binomial) %>%summary()

tstsol
tstroh
tstfb
```

Risk adjusted regression (TESTED) - blockPopulation and Preds
```{r}
rartst_sol = glm(tested ~ speakOtherLanguagesPropBG + preds + blockPopulation, data = rarDF, family = binomial) %>%summary()
rartst_roh = glm(tested ~ pRenterOccupiedHousesBG + preds + blockPopulation, data = rarDF, family = binomial) %>%summary()
rartst_fb = glm(tested ~ foreignBornPropBG + preds + blockPopulation, data = rarDF, family = binomial) %>%summary()

rartst_sol
rartst_roh
rartst_fb
```

Week 8+ 9:


Ben Huynh Github - Creating Table + Function for GLMResults
```{r}
getGLMResults <- function(mod) {
  c <- mod$coefficients[2]
  std.err <- mod$coefficients[2,2]
  CI <- (exp(std.err*qnorm(0.975))-1)*10
  cExp <- (exp(c)-1)*10 #interpret as per 10% increase in pop, not 100%
  upper <- cExp + CI
  lower <- cExp - CI
  pValue <- mod$coefficients[2,4]
  rr <- sqrt(exp(c))
  if(rr >= 1) {
    EValue <- rr + sqrt(rr*(rr-1))
  } else {
    EValue <- (1/rr) + sqrt((1/rr)*((1/rr)-1))
  }
  return(round(c(c,std.err,cExp,lower,upper,pValue,EValue),2))
}

generateRegressionTable <- function(rarSolResults, rarRohResults, rarFBResults,
                                    tstSolResults = NULL, tstRohResults = NULL, tstFbResults = NULL) {
  
  rarTable <- as.data.frame(rbind(rarSolResults, rarRohResults, rarFBResults))
  colnames(rarTable) <- c("Coefficient", "Std. err", "Increase", "Lower", "Upper", "p", "E")
  
  rarTable$Increase <- paste0(rarTable$Increase, " (", rarTable$Lower, ",", rarTable$Upper, ")")
  rarTable$Lower <- NULL
  rarTable$Upper <- NULL
  
  outcomeTable <- data.frame()
  
  if (!is.null(tstSolResults) && !is.null(tstRohResults) && !is.null(tstFbResults)) {
    outcomeTable <- as.data.frame(rbind(tstSolResults, tstRohResults, tstFbResults))
    colnames(outcomeTable) <- c("Coefficient_O", "Std. err_O", "Increase_O", "Lower_O", "Upper_O", "p_O", "E_O")
    
    outcomeTable$Increase_O <- paste0(outcomeTable$Increase_O, " (", outcomeTable$Lower_O, ",", outcomeTable$Upper_O, ")")
    outcomeTable$Lower_O <- NULL
    outcomeTable$Upper_O <- NULL
  }
    if (nrow(outcomeTable) > 0) {
    regTable <- cbind(rarTable, outcomeTable)
  } else {
    regTable <- rarTable
  }
  
  regTable$p <- ifelse(regTable$p == 0, "<0.001", regTable$p)
  if ("p_O" %in% colnames(regTable)) {
    regTable$p_O <- ifelse(regTable$p_O == 0, "<0.001", regTable$p_O)
  }
  
  return(regTable)
}

```

Mean Outcome results (model with blockPop)
```{r}
model_solResults <- getGLMResults(model_sol)
model_rohResults <- getGLMResults(model_roh)
model_fbResults <- getGLMResults(model_fb)
```

Tested results (model with blockPop and RAR with Preds+BlockPop)
```{r}

tstsolResults <- getGLMResults(tstsol)
tstrohResults <- getGLMResults(tstroh)
tstfbResults <- getGLMResults(tstfb)

rartst_solResults <- getGLMResults(rartst_sol)
rartst_rohResults <- getGLMResults(rartst_roh)
rartst_fbResults <- getGLMResults(rartst_fb)

```

**Tables**

Mean outcome
```{r}

regTable_mout = generateRegressionTable(model_solResults, model_rohResults, model_fbResults)
rownames(regTable_mout) = c('Speak Other Languages', "Renter Occupied Housing", "Foreign Born")
regTable_mout
```

Tested - Risk Adjusted Regression

```{r}
regTable_rartst = generateRegressionTable(rartst_solResults, rartst_rohResults, rartst_fbResults)
rownames(regTable_rartst) = c('Speak Other Languages', "Renter Occupied Housing", "Foreign Born")
regTable_rartst
```

Tested - Only with adjustment variable

```{r}
regTable_tst = generateRegressionTable(tstsolResults, tstrohResults, tstfbResults)
rownames(regTable_tst) = c('Speak Other Languages', "Renter Occupied Housing", "Foreign Born")
regTable_tst
```
















